\sys: Light-touch Data-Parallel Shell Processing

### 2020-05-12 pldi/draft.tex
Turn all paragraph to heading

% \paragraph{Generation}
% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams

@@

% \paragraph{Preliminaries}
% \paragraph{POSIX, GNU Core-utils, and beyond}

\paragraph{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert an {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to be split.
 % map :: (a -> b) -> [a] -> [b]

@@

\paragraph{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity---re\-turn\-ing the same results for the same inputs---but maintain internal state across their entire pass.
The details of this state and its propagation during piecewise element processing affect their distributability characteristics.

@@

\paragraph{File-system Commands}
The third class contains commands that access the file-system~\dfs.
This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving. % is grafted upon the file-system.

@@

\paragraph{Side-effectful Commands}
Another large class of commands are ones that have side-effects to memory~\sid.
Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.

@@

\paragraph{Irreversibly Side-effectful Commands}
Finally, a few commands (about 0.4\%) have side-effects outside \unix---\eg by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
Such commands are not distributable---they often have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism that locks these resources.

\paragraph{Discussion}
There are a few technical details about \sys's design decisions worth noting.

@@

\paragraph{Command Inputs and Outputs}

@@

\paragraph{Command-line Options}
As described earlier~\sx{bg}, com\-mand-line options and environment
variables can affect the operation of a command.  Options are a
particularly popular way to control a command, directly affecting a
command's parallelizability classification.  That is, commands are
assigned to default classes, and options can change the class in which
they belong.  As a simple example, consider \ttt{cat}.  By default
\ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it
has to keep track of a counter and output the line number with each line.

@@

%% \paragraph{Command Inputs and Outputs}
%% A second concern captured by the AL is the order of inputs and outputs.
%% Commands generally read multiple input streams (and, at cases, write to multiple output streams).
%% The invariants between inputs and outputs must be maintained over the distributed execution of the program.
%% % but these maintain different invariants, enabling very different distributability characteristics.

@@

\paragraph{Parallelizability Annotation Language}

@@

\paragraph{Parallelizable Pure Commands}

@@

\paragraph{Edges -- Files}

@@

\paragraph{Nodes -- Commands}

@@

\paragraph{\todo{One-input-output} Commands}

@@

\paragraph{Stateless and Parallelizable Pure Commands}

@@

\paragraph{Parallelization Transformations}
%
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a dataflow graph is its output, this
optimization $ v(x_1.x_2...x_n, s) \Rightarrow v(x_1, s).v(x_2,
s)...v(x_n, s)$ can be shown to preserve the behaviour of the
graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-fold pair $(m, r)$ as $
v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
s)$. A visual representation of this transformation can be seen in
\Cref{fig:parallelization-transformation}.

@@

\paragraph{Parallelizable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Parallelizable regions are program sub-expressions that can be
parallelized safely without altering the total program behaviour.
%% respect to their sequential execution.
The search for these regions can be guided by the structure of shells
programs, which inherently contains information about components that
can be executed independently and barriers that are natural points for
enforcing synchronization. Consider the following example:

@@

\paragraph{Translation Pass}

@@

\paragraph{Pure Fold Function Implementations}

@@

\paragraph{Distributed Operating Systems and Shells}
A plethora of systems assist in the construction of distributed
software.  Distributed operating systems~\cite{ousterhout1988sprite,
  mullender1990amoeba, pike1990plan9, barak1998mosix} and programming
languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11} hide
many distribution problems, as long as a program uses the provided
abstractions---which are strongly coupled to the underlying operating
or runtime system.
% simplify many of the problems of distribution and

@@

\paragraph{Cluster Computing}
Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12,
  naiad:13} and domain-specific languages~\cite{alvaro2011consistency,
  distal:13, meiklejohn2015lasp} go a step further in terms of
automation by providing a few carefully-designed primitives.  These
primitives make strong assumptions about the nature of the
computation---\eg strongly-eventual commutative functions that can
proceed in parallel.  By targeting specific classes of computation
(\emph{viz.} \sys's distributability), they render distributed
computation efficiently.

@@

\paragraph{Annotation-based Distribution}
Recent systems on light-touch distribution, such as
Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general
than cluster-computing frameworks.  They also lower (but not entirely
eliminate) developer effort by automating most of the rewriting
process.
% Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
For the critical points where user input is required, they provide a
high-level declarative DSL with varying (but limited) degrees of
expressive power.
% Annotations are used by ---complicated by the use of third-party
% libraries similar to \sys's extensions.
Unfortunately, a challenge with this approach is that users composing
applications by using third-party libraries do not understand their
distributability details well-enough to annotate them.

@@

% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
\paragraph{Low-level Parallelization}
Instruction-level parallelization has a longer history, starting from
explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2}
and continuing with compilers that attempt to automatically extract
parallelism~\cite{padua1993polaris,hall1996maximizing}.  These systems
operate at a lower level than \sys (\eg that of instructions or loops
instead of pipe boundaries) and typically do not exploit runtime
information.

@@

\paragraph{Correct Distribution of Dataflow Graphs}
%
The dataflow graph is a prevalent model in several areas of data
processing, as it is used by several batch-~\cite{mapreduce:08,
  spark:12} and stream-processing systems~\cite{murray2013naiad,
  carbone2015flink}. Despite its popularity, most systems perform
optimizations that do not preserve its semantics, introducing subtle
erroneous behaviours. Recent work~\cite{HSSGG2014, SHGW2015,
  MSAIT2019} attempts to address this issue by performing
optimizations only in cases where correctness is preserved. \sys draws
inspiration from these efforts, as we are also interested in
optimizations that exploit parallelization but preserve the semantics
of the underlying model. However, our proposed dataflow model is
different from these other works, as nodes in our model read from
their inputs \emph{in sequence}. Due to this intricacy, previous work
is not straightforwardly applicable.

@@

\paragraph{Limitations}
Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.

@@

\paragraph{Future Work}
% \label{limitation}
% \item No cycles (multiple commands writing and reading from the same file)
There are several worthwhile directions for future work.  One
direction would be an extension of the distributability analysis of
\sys, in order to guide it using dynamic information. This could
enable a formal study of the analysis~\sx{impl} to show that it
returns regions that are safe to distribute.  Recent work on
formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
such an attempt possible.

@@

\paragraph{Conclusion}
This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
\sys's insight is that shell pipelines already express streaming computations that can be automatically distributed.
To achieve its goal, \sys
  decomposes primitives into distributability classes,
  identifies high-distributability stages,
  applies rewriting rules for largest possible subprograms,
  and orchestrates the execution of the distributed program.
% Its runtime component provides orchestration and planning support during the execution of the program.
Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.

>>>>

% \heading{Generation}
% Expansion---the key is that expansion and evaluation are handled by the shell at an earlier stage 
% The most powerful---and interesting, for \ttt{dish}---type of expansion is subshell expansion, that allows replacing entire input streams

@@

% \heading{Preliminaries}
% \heading{POSIX, GNU Core-utils, and beyond}

\heading{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands that can be expressed as a purely functional $map$ or $filter$---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert an {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness, nor caution about where to be split.
 % map :: (a -> b) -> [a] -> [b]

@@

\heading{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity---re\-turn\-ing the same results for the same inputs---but maintain internal state across their entire pass.
The details of this state and its propagation during piecewise element processing affect their distributability characteristics.

@@

\heading{File-system Commands}
The third class contains commands that access the file-system~\dfs.
This is the largest class, as the file (and file-system) abstraction is central to the \unix design and philosophy~\cite{unix}.
In fact, \unix uses the file-system as a proxy to several file-unrelated operations---\eg access control and device driving. % is grafted upon the file-system.

@@

\heading{Side-effectful Commands}
Another large class of commands are ones that have side-effects to memory~\sid.
Because of timesharing and multi-tasking---\ie the fact that many users can operate multiple terminals on single \unix machine---a subset of these commands are devoted to accessing and sharing memory such as environment variables and shell parsing flags.

@@

\heading{Irreversibly Side-effectful Commands}
Finally, a few commands (about 0.4\%) have side-effects outside \unix---\eg by printing a file with \ttt{lpr} or restarting with \ttt{reboot}.
Such commands are not distributable---they often have to refer to a single location---and have to be guarded with pessimistic transaction control mechanism that locks these resources.

\heading{Discussion}
There are a few technical details about \sys's design decisions worth noting.

@@

\heading{Command Inputs and Outputs}

@@

\heading{Command-line Options}
As described earlier~\sx{bg}, com\-mand-line options and environment
variables can affect the operation of a command.  Options are a
particularly popular way to control a command, directly affecting a
command's parallelizability classification.  That is, commands are
assigned to default classes, and options can change the class in which
they belong.  As a simple example, consider \ttt{cat}.  By default
\ttt{cat} is in \sta, but with \ttt{-n} it jumps into \pur because it
has to keep track of a counter and output the line number with each line.

@@

%% \heading{Command Inputs and Outputs}
%% A second concern captured by the AL is the order of inputs and outputs.
%% Commands generally read multiple input streams (and, at cases, write to multiple output streams).
%% The invariants between inputs and outputs must be maintained over the distributed execution of the program.
%% % but these maintain different invariants, enabling very different distributability characteristics.

@@

\heading{Parallelizability Annotation Language}

@@

\heading{Parallelizable Pure Commands}

@@

\heading{Edges -- Files}

@@

\heading{Nodes -- Commands}

@@

\heading{\todo{One-input-output} Commands}

@@

\heading{Stateless and Parallelizable Pure Commands}

@@

\heading{Parallelization Transformations}
%
Based on these equations, we can define a \emph{node parallelization
  transformation} on a stateless node $v$ that is preceded by a
\ttt{cat} with $n$ input streams and is followed by a node $v'$. The
transformation replaces $v$ with $n$ new nodes, routing each of the
$n$ input streams to one of them, and commutes the \ttt{cat} node
after them to concatenate their outputs and transfer them to
$v'$. Since each incoming edge represents a stream of data $x_i :
D^*$, and the only behaviour of a dataflow graph is its output, this
optimization $ v(x_1.x_2...x_n, s) \Rightarrow v(x_1, s).v(x_2,
s)...v(x_n, s)$ can be shown to preserve the behaviour of the
graph. Furthermore, it can be straightforwardly extended to pure
commands that can be implemented by a map-fold pair $(m, r)$ as $
v(x_1.x_2...x_n, s) \Rightarrow r(m(x_1, s).m(x_2, s)...m(x_n, s),
s)$. A visual representation of this transformation can be seen in
\Cref{fig:parallelization-transformation}.

@@

\heading{Parallelizable Regions}
%
%% \kk{The goal of this subsection is to explain that Dish does a safe
%%   but effective analysis to identify which parts of the scripts to
%%   parallelize.}
%
Parallelizable regions are program sub-expressions that can be
parallelized safely without altering the total program behaviour.
%% respect to their sequential execution.
The search for these regions can be guided by the structure of shells
programs, which inherently contains information about components that
can be executed independently and barriers that are natural points for
enforcing synchronization. Consider the following example:

@@

\heading{Translation Pass}

@@

\heading{Pure Fold Function Implementations}

@@

\heading{Distributed Operating Systems and Shells}
A plethora of systems assist in the construction of distributed
software.  Distributed operating systems~\cite{ousterhout1988sprite,
  mullender1990amoeba, pike1990plan9, barak1998mosix} and programming
languages~\cite{erlang:96, acute:05, mace:07, cloudhaskell:11} hide
many distribution problems, as long as a program uses the provided
abstractions---which are strongly coupled to the underlying operating
or runtime system.
% simplify many of the problems of distribution and

@@

\heading{Cluster Computing}
Modern analytics frameworks~\cite{mapreduce:08, ciel:11, spark:12,
  naiad:13} and domain-specific languages~\cite{alvaro2011consistency,
  distal:13, meiklejohn2015lasp} go a step further in terms of
automation by providing a few carefully-designed primitives.  These
primitives make strong assumptions about the nature of the
computation---\eg strongly-eventual commutative functions that can
proceed in parallel.  By targeting specific classes of computation
(\emph{viz.} \sys's distributability), they render distributed
computation efficiently.

@@

\heading{Annotation-based Distribution}
Recent systems on light-touch distribution, such as
Ignis~\cite{ignis:19} and Mozart~\cite{mozart:19}, are more general
than cluster-computing frameworks.  They also lower (but not entirely
eliminate) developer effort by automating most of the rewriting
process.
% Both work on high-level, managed languages---server-side JavaScript for Ignis and Python for Mozart---that allow runtime transformations without affecting the broader environment under which a program executes.
For the critical points where user input is required, they provide a
high-level declarative DSL with varying (but limited) degrees of
expressive power.
% Annotations are used by ---complicated by the use of third-party
% libraries similar to \sys's extensions.
Unfortunately, a challenge with this approach is that users composing
applications by using third-party libraries do not understand their
distributability details well-enough to annotate them.

@@

% Nice intro: http://homepages.inf.ed.ac.uk/bfranke/Publications/pldi121-tournavitis.pdf
\heading{Low-level Parallelization}
Instruction-level parallelization has a longer history, starting from
explicit \ttt{DOALL} and \ttt{DOACROSS} annotations~\cite{par1, par2}
and continuing with compilers that attempt to automatically extract
parallelism~\cite{padua1993polaris,hall1996maximizing}.  These systems
operate at a lower level than \sys (\eg that of instructions or loops
instead of pipe boundaries) and typically do not exploit runtime
information.

@@

\heading{Correct Distribution of Dataflow Graphs}
%
The dataflow graph is a prevalent model in several areas of data
processing, as it is used by several batch-~\cite{mapreduce:08,
  spark:12} and stream-processing systems~\cite{murray2013naiad,
  carbone2015flink}. Despite its popularity, most systems perform
optimizations that do not preserve its semantics, introducing subtle
erroneous behaviours. Recent work~\cite{HSSGG2014, SHGW2015,
  MSAIT2019} attempts to address this issue by performing
optimizations only in cases where correctness is preserved. \sys draws
inspiration from these efforts, as we are also interested in
optimizations that exploit parallelization but preserve the semantics
of the underlying model. However, our proposed dataflow model is
different from these other works, as nodes in our model read from
their inputs \emph{in sequence}. Due to this intricacy, previous work
is not straightforwardly applicable.

@@

\heading{Limitations}
Our implementation is limited in many ways, so as to succeed in proving the key hypothesis---that scaling out shell pipelines can be automated and correct with respect to some assumptions.

@@

\heading{Future Work}
% \label{limitation}
% \item No cycles (multiple commands writing and reading from the same file)
There are several worthwhile directions for future work.  One
direction would be an extension of the distributability analysis of
\sys, in order to guide it using dynamic information. This could
enable a formal study of the analysis~\sx{impl} to show that it
returns regions that are safe to distribute.  Recent work on
formalizing the semantics of the POSIX shell~\cite{smoosh:20} make
such an attempt possible.

@@

\heading{Conclusion}
This paper presented \sys, a shell variant that automatically and correctly scales out distribution-oblivious shell pipelines. 
\sys's insight is that shell pipelines already express streaming computations that can be automatically distributed.
To achieve its goal, \sys
  decomposes primitives into distributability classes,
  identifies high-distributability stages,
  applies rewriting rules for largest possible subprograms,
  and orchestrates the execution of the distributed program.
% Its runtime component provides orchestration and planning support during the execution of the program.
Experiments with complex pipelines show substantial speedups and the ability to operate on large input datasets, all without any developer input.
